---
title: "class 8 halloween"
format: pdf
---

```{r}
candy_file <- read.csv("candy-data.csv", row.names = 1)

head(candy_file)
```

Q1. How many different candy types are in this dataset?

```{r}
nrow(candy_file)
```


Q2. How many fruity candy types are in the dataset?
The functions dim(), nrow(), table() and sum() may be useful for answering the first 2 questions.


```{r}
dim(candy_file)
```

```{r}
sum(candy_file$fruity)
```
Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy_file["Almond Joy",]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy_file["Kit Kat",]$winpercent
```


Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?
```{r}
candy_file["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(skimr)
skim(candy_file)
```
```{r}
skimr::skim(candy_file)
```

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?


hist

Q7. What do you think a zero and one represent for the candy$chocolate column?
0 - no chocolate
1- chocolate

```{r}
candy_file$chocolate
choc.ind <- as.logical(candy_file$chocolate)
fruit.ind <- as.logical(candy_file$fruity)
choc.win <- candy_file[choc.ind,]$winpercent
choc.win

```

```{r}
mean(choc.ind)
mean(fruit.ind)

mean(candy_file$chocolate)
mean(candy_file$fruity)
```

Q8. Plot a histogram of winpercent values

```{r}
hist(candy_file$winpercent)
```

Q9. Is the distribution of winpercent values symmetrical?
NO



Q10. Is the center of the distribution above or below 50%?

```{r}
candy_file$winpercent[as.logical(candy_file$nougat)]
```

Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy_file$chocolate) >= mean(candy_file$fruity) 
```





Q12. Is this difference statistically significant?

## 
##  Welch Two Sample t-test
## 
## data:  chocolate and fruity
## t = 6.2582, df = 68.882, p-value = 2.871e-08
## alternative hypothesis: true difference in means is not equal to 0
## 95 percent confidence interval:
##  11.44563 22.15795
## sample estimates:
## mean of x mean of y 
##  60.92153  44.11974


Q13. What are the five least liked candy types in this set?

```{r}
x <- c(5,2,3,6)
sort(x)
sort(x, decreasing=TRUE)
sort(x, decreasing =FALSE)
```

```{r}
x
order(x)
```

```{r}
x[order(x)]
```

```{r}
y<-c("D", "A", "E")
order(y)
```


Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
order.win <- order(candy_file$winpercent)
candy_file[order.win[1:5],1]
tail(order.win)
head(order.win)
```
```{r}
head(candy_file[order(candy_file$winpercent),], n=5)
tail(candy_file[order(candy_file$winpercent),], n=5)
```

```{r}
x <- order(candy_file$winpercent)
order(candy_file$winpercent)
```


Q15. Make a first barplot of candy ranking based on winpercent values.
```{r}
library(ggplot2)
ggplot(candy_file) +
  aes(winpercent, rownames(candy_file))+
  geom_col()
```
Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy_file) +
  aes(winpercent, rownames(candy_file), reorder(candy_file, FUN=mean(candy_file$winpercent)))+
  geom_col()
```


```{r}
mycols <- "gray"
ggplot(candy_file) + 
  aes(winpercent, reorder(rownames(candy_file),winpercent)) +
  geom_col(bg=mycols)
```

```{r}
mycols <- rep("gray",nrow(candy_file))
           
ggplot(candy_file) + 
  aes(winpercent, reorder(rownames(candy_file),winpercent)) +
  geom_col(bg=mycols)
```

```{r}
mycols <- rep("gray", nrow(candy_file))
mycols[as.logical(candy_file$chocolate)] <- "brown"
mycols=rep("black", nrow(candy_file))
mycols[as.logical(candy_file$chocolate)] = "chocolate"
mycols[as.logical(candy_file$bar)] = "orange"
mycols[as.logical(candy_file$fruity)] = "pink"
mycols
ggplot(candy_file) + 
  aes(winpercent, reorder(rownames(candy_file),winpercent)) +
  geom_col(bg=mycols)
```
 Q17. What is the worst ranked chocolate candy?
 Resse's Peanut Butter cup
 
 Q18. What is the best ranked fruity candy?
 Nik L Nip
 
# How about a plot of price vs win




Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
library(ggrepel)

ggplot(candy_file) +
  aes(winpercent, pricepercent, label=rownames(candy_file)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```



Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy_file$pricepercent, decreasing = TRUE)
head( candy_file[ord,c(11,12)], n=5 )
```


Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy_file) +
  aes(pricepercent, reorder(rownames(candy_file), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy_file), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
```{r}
library(corrplot)
cij <- cor(candy_file)
corrplot(cij)
```
chocolate and fruity


Q23. Similarly, what two variables are most positively correlated?
chocolate and bar abd oricepercent and winpercent

#Principal Component Analysis

```{r}
pca <- prcomp(candy_file, scale=TRUE)
summary(pca)
plot(pca$x[,1:2], col=mycols, pch=16)
```
```{r}
my_data <- cbind(candy_file, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


